Master's Project Journal

### 2019-04-11 thesis/FPGABasedADPLLs.tex
more pngs and svgs

\section{Chapter Overview}
The first step in creating an \ac{FPGA} based network of \ac{ADPLL}s is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of an \ac{FPGA} necessitates a number of compromises in the design of a given block which limits transferability to \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in an \ac{FPGA} based \ac{ADPLL} made. A number of blocks have implement purely digital circuitry and as such can be transferred in their entirety from an \ac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail.

@@

\subsection{\acs{FPGA} Driven, Linear Period \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc1}
	\caption{\acs{FPGA} Driven, Linear Period \acs{DCO} \ac{RTL} Diagram.}
	\label{fig:osc1}
\end{figure}
The first design of \ac{DCO} to be examined is of the type used by Zianbetov in his \ac{ADPLL} network test bed and relies on a counter driven by the clock manager on the \ac{FPGA} \cite{zianbetov2013phd}. At each event on the \ac{FPGA} provided clock a counter is incremented, overflowing upon increment past the maximum possible value. The \ac{MSB} of this counter forms the waveform generated by this oscillator, the period of which is controlled through an adjustable value that is loaded into the counter once overflow is reached, and forms the starting points for the counter. The period of oscillation is given by:
\begin{equation}
	T_{osc} = \big(2^{width} -~(BIAS+CC)~\big)\times T_{FPGA}
\end{equation}
where $T_{FPGA}$ is the clock period of the \ac{FPGA}, $CC$ is the control code input, $BIAS$ centres the oscillator in the middle of the tuning range in the event that the control code is zero and $width$ is the width of the counter used. As the only variable here is the control code, period step of this design is $T_{FPGA}$, thereby providing period linearity with respect to the control code. This is the key advantage of this design, as most \ac{ASIC} implementations of a \ac{DCO} are also linear in period. The other main reason to choose this design is that its \ac{FPGA} clocked nature allows for exact control over the frequency of operation, and the number of tunable parameters make it possible to configure multiple ways to achieve the same frequencies of operation. Combined these attributes make it very easy to create an oscillator that emulates the behaviour of a design intended for an \ac{ASIC}, however, at a greatly reduced frequency. This restriction on the frequency of operation arises out of the period step size, which in order to obtain a good resolution must be orders of magnitude smaller than the intended period to be generated. As the output waveform is taken from the counter's \ac{MSB}, the reload value of the counter must never go beyond $2^{width-1}$, as otherwise the output waveform will become a constant 1. As the reload value varies the low time of the \ac{MSB}, if the desired output waveform is a square wave this design will not be suitable.

@@

\subsection{\acs{FPGA} Driven, Linear Frequency \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc2}
	\caption{\acs{FPGA} Driven, Linear Period \acs{DCO} \ac{RTL} Diagram.}
	\label{fig:osc2}
\end{figure}
The second \ac{FPGA} clocked oscillator is similar in most attributes to the above design but eschews period linearity for frequency linearity. Again the overflow property of a counter is used with the counter's \ac{MSB} as output of the block, however, this time it forms a square wave. Rather than setting the reload value of the counter, instead the increment is adjusted depending on the control code, thus requiring $\frac{2^N}{BIAS+CC}$ increments to overflow. Accordingly the frequency of operation is set by:
\begin{equation}
	f_{osc} = f_{FPGA}\times\frac{BIAS+CC}{2^{width}}
\end{equation}
Here the control code $CC$ and bias are added to the value stored in the accumulator at each event of the \ac{FPGA}, clock until overflow is reached. This occurs at $2^{width}$ where, as before, $width$ is the bit width of the counter, thus valuing each control code increment at $\frac{f_{FPGA}}{2^{width}}$ Hz. As with the previous design, this oscillator is better suited to frequencies where the output of the \ac{DCO} is orders of magnitude lower than the clock signal driving it, as this ensures that the incremental change due to the control code remains a small fraction of the period.

@@

\subsection{Ring Oscillator}
The best approximation of the non-idealities of an \ac{ASIC} \ac{DCO} implementation can be obtained using a \ac{RO}, also know as an inverter ring, as the \ac{DCO}. In an \ac{ASIC} an \ac{RO} is constructed by connecting a number of NOT gates back in a chain, with the output of the last NOT gate fed input the input of the first. As NOT gates are not available on an \ac{FPGA}, they are replaced by inverter primitives. The inverter ring is an inherently unstable circuit that leverages the propagation delay through an odd number of inverters to create a waveform that, when viewed at a fixed position, is a square wave. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.4\textwidth]{../RO-wiki}
	\caption{Basic Ring Oscillator \ac{RTL} Diagram \cite{ro_wiki}.}
	\label{fig:ro_wiki}
\end{figure}

@@

\section{Feedback Divider}
Many \ac{PLL}s, be they digital or analogue, incorporate a divider in the feedback path so that comparison can be made with a reference operating at a lower frequency. This is a feature that is advantageous for \ac{ADPLL} networks, as comparison using this divided clock allows the synchronisation signal sent between modules to be of a frequency orders of magnitude lower than that of the clock distribution network and thus require circuitry that consumes significantly less energy. This division can be obtained trivially on an \ac{FPGA} by a number of means, of which this thesis will mention just two. The first of these methods allows for integer division, by inverting a signal each time a counter reaches a certain value, $k$. When this condition is true the value stored in the counter is also reset to zero, thus implementing division of the generated clock frequency by $2k$.
\begin{figure}[h]%
	\centering
	\subfloat[Divider 1]{{\includegraphics[width=0.6\textwidth]{../divider1} }}\\
	\subfloat[Divider 2]{{\includegraphics[width=0.6\textwidth]{../divider2} }}%
	\caption{Divider \ac{RTL} Diagrams.}
	\label{fig:divs}
\end{figure}
By limiting to division by powers of two, the divider can be simplified with the removal of any comparison logic. One flip-flop and an inverter are required per power of two. As in Figure \ref{fig:divider_2}, the inverted $\overline{Q}$ output of each flip-flop is inverted and connected to the $D$ input, with the non-inverted $Q$ of the previous flip-flop acting as the clock. The first flip-flop in the chain is clocked by the \ac{DCO} output.

@@

Accordingly any intended error combiner design implementing an average should also be reconfigurable to satisfy the reconfigurability constraint. Given this is a digital system the most straightforward way to implement this is as a weighted sum, followed by a division. The final \ac{ADPLL} network will be laid out in a Cartesian grid so a maximum of four values will possibly be combined, with only the non-corner edge nodes having a number of neighbours that is not a power of two.
The simplest, and most processing time efficient, method of calculating the weighted sum restricts the weights to powers of two, as this permits their implementation as a left bit shift which is trivially achieved in hardware by a part select.
Similarly this reasoning can be expanded to the division, where the complicated division by three for a non-corner edge node is instead replaced by division by four. This means the division is implemented identically by all error combiners in the network, and implemented trivially by a right shift of two bits.%TODO ask pierre re 3 div 4

>>>>

\section{Chapter Overview}
The first step in creating an \ac{FPGA} based network of \ac{ADPLL}s is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of an \ac{FPGA} necessitates a number of compromises in the design of a given block which limits transferability to \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in an \ac{FPGA} based \ac{ADPLL} made. A number of blocks have implement purely digital circuitry and as such can be transferred in their entirety from an \ac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail.
\begin{figure}[h] %TODO redo this image
	\centering
	\includegraphics[width=\textwidth]{../network_adpll}
	\caption{\ac{ADPLL} designed for use in a Cartesian grid network.}
	\label{fig:dist_adpll}
\end{figure}

@@

\subsection{\acs{FPGA} Driven, Linear Period \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc1}
	\caption{\acs{FPGA} driven, linear period \acs{DCO} \ac{RTL} Diagram.}
	\label{fig:osc1}
\end{figure}
The first design of \ac{DCO} to be examined is of the type used by Zianbetov in his \ac{ADPLL} network test bed and relies on a counter driven by the clock manager on the \ac{FPGA} \cite{zianbetov2013phd}. At each event on the \ac{FPGA} provided clock a counter is incremented, overflowing upon increment past the maximum possible value. The \ac{MSB} of this counter forms the waveform generated by this oscillator, the period of which is controlled through an adjustable value that is loaded into the counter once overflow is reached, and forms the starting points for the counter. The period of oscillation is given by:
\begin{equation}
	T_{osc} = \big(2^{width} -~(BIAS+CC)~\big)\times T_{FPGA}
\end{equation}
where $T_{FPGA}$ is the clock period of the \ac{FPGA}, $CC$ is the control code input, $BIAS$ centres the oscillator in the middle of the tuning range in the event that the control code is zero and $width$ is the width of the counter used. As the only variable here is the control code, period step of this design is $T_{FPGA}$, thereby providing period linearity with respect to the control code. This is the key advantage of this design, as most \ac{ASIC} implementations of a \ac{DCO} are also linear in period. The other main reason to choose this design is that its \ac{FPGA} clocked nature allows for exact control over the frequency of operation, and the number of tunable parameters make it possible to configure multiple ways to achieve the same frequencies of operation. Combined these attributes make it very easy to create an oscillator that emulates the behaviour of a design intended for an \ac{ASIC}, however, at a greatly reduced frequency. This restriction on the frequency of operation arises out of the period step size, which in order to obtain a good resolution must be orders of magnitude smaller than the intended period to be generated. As the output waveform is taken from the counter's \ac{MSB}, the reload value of the counter must never go beyond $2^{width-1}$, as otherwise the output waveform will become a constant 1. As the reload value varies the low time of the \ac{MSB}, if the desired output waveform is a square wave this design will not be suitable.

@@

\subsection{\acs{FPGA} Driven, Linear Frequency \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc2}
	\caption{\acs{FPGA} driven, linear frequency \acs{DCO} \ac{RTL} Diagram.}
	\label{fig:osc2}
\end{figure}
The second \ac{FPGA} clocked oscillator is similar in most attributes to the above design but eschews period linearity for frequency linearity. Again the overflow property of a counter is used with the counter's \ac{MSB} as output of the block, however, this time it forms a square wave. Rather than setting the reload value of the counter, instead the increment is adjusted depending on the control code, thus requiring $\frac{2^N}{BIAS+CC}$ increments to overflow. Accordingly the frequency of operation is set by:
\begin{equation}
	f_{osc} = f_{FPGA}\times\frac{BIAS+CC}{2^{width}}
\end{equation}
Here the control code $CC$ and bias are added to the value stored in the accumulator at each event of the \ac{FPGA}, clock until overflow is reached. This occurs at $2^{width}$ where, as before, $width$ is the bit width of the counter, thus valuing each control code increment at $\frac{f_{FPGA}}{2^{width}}$ Hz. As with the previous design, this oscillator is better suited to frequencies where the output of the \ac{DCO} is orders of magnitude lower than the clock signal driving it, as this ensures that the incremental change due to the control code remains a small fraction of the period.

@@

\subsection{Ring Oscillator}
The best approximation of the non-idealities of an \ac{ASIC} \ac{DCO} implementation can be obtained using a \ac{RO}, also know as an inverter ring, as the \ac{DCO}. In an \ac{ASIC} an \ac{RO} is constructed by connecting a number of NOT gates back in a chain, with the output of the last NOT gate fed input the input of the first. As NOT gates are not available on an \ac{FPGA}, they are replaced by inverter primitives. The inverter ring is an inherently unstable circuit that leverages the propagation delay through an odd number of inverters to create a waveform that, when viewed at a fixed position, is a square wave. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.4\textwidth]{../RO-wiki}
	\caption[Basic ring oscillator \ac{RTL} diagram]{Basic ring oscillator \ac{RTL} diagram \cite{ro_wiki}.}
	\label{fig:ro_wiki}
\end{figure}

@@

\section{Frequency Divider}
Many \ac{PLL}s, be they digital or analogue, incorporate a frequency divider in the feedback path so that comparison can be made with a reference operating at a lower frequency. This is a feature that is advantageous for \ac{ADPLL} networks, as comparison using this divided clock allows the synchronisation signal sent between modules to be of a frequency orders of magnitude lower than that of the clock distribution network and thus require circuitry that consumes significantly less energy. This division can be obtained trivially on an \ac{FPGA} by a number of means, of which this thesis will mention just two. The first of these methods allows for integer division, by inverting a signal each time a counter reaches a certain value, $k$. When this condition is true the value stored in the counter is also reset to zero, thus implementing division of the generated clock frequency by $2k$.
\begin{figure}[h]%
	\centering
	\subfloat[Divider 1]{{\includegraphics[width=0.6\textwidth]{../divider1} }}\\
	\subfloat[Divider 2]{{\includegraphics[width=0.6\textwidth]{../divider2} }}%
	\caption{Frequency divider \ac{RTL} diagrams.}
	\label{fig:divs}
\end{figure}
By limiting to division by powers of two, the divider can be simplified with the removal of any comparison logic. One flip-flop and an inverter are required per power of two. As in Figure \ref{fig:divider_2}, the inverted $\overline{Q}$ output of each flip-flop is inverted and connected to the $D$ input, with the non-inverted $Q$ of the previous flip-flop acting as the clock. The first flip-flop in the chain is clocked by the \ac{DCO} output.

@@

Accordingly, any intended error combiner design implementing an average should also be reconfigurable, so that it satisfies the reconfigurability constraint. Given this is a digital system the most straightforward way to implement this is as a weighted sum, followed by a division. The final \ac{ADPLL} network will be laid out in a Cartesian grid so a maximum of four values will possibly be combined, with only the non-corner edge nodes having a number of neighbours that is not a power of two.
The simplest, and most processing time efficient, method of calculating the weighted sum restricts the weights to powers of two, as this permits their implementation as a left bit shift which is trivially achieved in hardware by a part select.
Similarly this reasoning can be expanded to the division, where the complicated division by three for a non-corner edge node is instead replaced by division by four. This means the division is implemented identically by all error combiners in the network, and implemented trivially by a right shift of two bits.%TODO ask pierre re 3 div 4
